Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-04-11 analysis/paper/paper.Rmd
incorporated MH edits

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is within a section of the Salmon River containing historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

@@

In additional to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including both natural-origin juveniles and hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that juvenile emigration and adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide for recreational fishing opportunities in the Upper Salmon River. Consequently, reducing predation mortality in Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population. Predation on juveniles from any of the ESA-listed salmonid species is likely detrimental to their recovery.

@@

```{r map, out.width = "85%", fig.cap = "Map of the Deadwater Slough study area within the Upper Salmon River MPG. The Dump Creek alluvial fan is loacted at the downstream end of the study reach."}
knitr::include_graphics("../figures/Study Area Map 5.png")

@@

```{r juv-con-p, fig.cap = "Estimated number of juvenile Chinook Salmon consumed by Northern Pikeminnow from a sensitivity analysis of variable diet scenarios. The proportion of Chinook Salmon in the total fish prey consumed by Northern Pikminnow were evaluated for 30%, 50%, and 65% scenarios, similar to those reported by Zimmerman and Ward 1999."}

>>>>

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is within a section of the Salmon River containing historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

@@

In additional to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon (both natural-origin and hatchery releases) are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that juvenile emigration and adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide for recreational fishing opportunities in the Upper Salmon River. Consequently, reducing predation mortality in Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population. Predation on juveniles from any of the ESA-listed salmonid species is likely detrimental to their recovery.

@@

```{r map, out.width = "85%", fig.cap = "Map of the Deadwater Slough study area within the Upper Salmon River MPG. The Dump Creek alluvial fan is located at the downstream end of the study reach."}
knitr::include_graphics("../figures/Study Area Map 5.png")

@@

```{r juv-con-p, fig.cap = "Estimated number of juvenile Chinook Salmon consumed by Northern Pikeminnow from a sensitivity analysis of variable diet scenarios. The proportion of Chinook Salmon in the total fish prey consumed by Northern Pikeminnow were evaluated for 30%, 50%, and 65% scenarios, similar to those reported by Zimmerman and Ward 1999."}
